Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-08-22 write-up/covid-writeup.tex
Update covid-writeup.tex
Fixed all visualizations to grayscale

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  We build on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  In particular, we quantify the interaction between measurement-error and selection bias.  The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.

@@

@@

\subsection{Application of SEIR model}
\label{section:modelbased}

Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates. Based on published age-specific IFRs~\cite{Levin2020}, Figure~\ref{fig:logifr} shows that IFR seems to closely follow a log-linear relationship with age. Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
\begin{table}[!th]
\begin{tabular}{c | c c c c c c c c}
Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
\end{tabular}
\caption{IFR by age-strata}
\label{tab:ifrperage}
\end{table}
The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.

The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.  Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.  Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
$$
\beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
$$
where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.  Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$. See Section~\ref{app:modelbased} in the supplementary materials for details on prior specification.

@@

\bibliographystyle{plain}
\bibliography{covid-refs}

@@

For binary outcomes, we have
\begin{equation} \label{eq:binaryrho}
\rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
\end{equation}

@@

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig1.png}
 \caption{Symptom Likelihood Given Negative Test}
 \label{fig:symptomlik1_model2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig2.png}
 \caption{Symptom Likelihood Given Positive Test}
 \label{fig:symptomlik2_model2}
\end{subfigure}
\caption{Likelihood of COVID-19 contact}
\label{fig:symptomlik}
\end{figure}

@@

\section{Model-based: Prior specification}
\label{app:modelbased}

>>>>

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and population size.  We build on the work of \cite{Meng2018} who studied an error decomposition to understand the relationship between selection bias and population size.  In particular, we quantify the interaction between measurement-error and selection bias.  The sign and magnitude of the statistical decomposition can change drastically; the impact on the effective sample size can therefore be quite large.
\textcolor{red}{Acknowledge selection bias and measurement error.}

@@

\begin{rmk}[Reporting Delays]

\end{rmk}

@@

\subsection{Application of SEIR model}
\label{section:application_of_modelbased}

Here we consider the multinomial SEIR model using observed death data as presented in Figure~\ref{fig:in-deaths}. The model, presented in Section~\ref{section:modelbased}, requires infection fatality rates to be specified per-strata.  In this paper, similar to~\cite{Johndrow2020}, we specify fixed per-strata infection fatality rates. Based on published age-specific IFRs~\cite{Levin2020}, IFR closely follows a log-linear relationship with age. Combining across published age-specific IFRs, Table~\ref{tab:ifrperage} presents estimates of the average IFR within each strata using the age-distribution.
\begin{table}[!th]
\begin{tabular}{c | c c c c c c c c}
Age Strata & 0-19 & 20-29 & 30-39 & 40-49 & 50-59 & 60-69 & 70-79 & 80+ \\ \hline
IFR & 0.003\% & 0.014\% & 0.046\% & 0.144\% & 0.456\% & 1.443\% & 4.567\% & 15.468\%
\end{tabular}
\caption{IFR by age-strata}
\label{tab:ifrperage}
\end{table}
The discrete-time distribution~$\{ \theta_{j} \}$ is a discretized version of a truncated normal distribution with mean $25$, standard deviation~$5$, minimum value $0$, and maximum value $44$.  This was chosen to closely mimic the distribution from~\cite{Johndrow2020}.

The SEIR model in equation~\eqref{eq:seir} is extended to allow for time-varying transmission rates~$\beta_t$ to account for public policy changes discussed in Section~\ref{section:publicpolicyindiana} over the time period considered.  Here, we insert three change points $(t_1,t_2,t_3)$ on March 23rd, June 15th, and October 1st of 2020 respectively.  Given sensitivity to these choices, instead of considering a multiplicative structure, we allow for delayed implementation and slow change in the transmission rate by using weights:
$$
\beta_t = \sum_{k=1}^{3} \left(\beta_{k-1} (1-w^{(k)}_t ) + \beta_{k} w_t\right),\quad w^{(k)}_t = (1 + \exp \left( -\xi (t - t_k - \nu) \right))^{-1}
$$
where $\nu \geq 0$ is the potential delay and $\xi$ is the rate of change.  Each weight~$w_t^{(k)} = 0$ for $t < t_k$ and $w_t^{(k)} \to 1$ as $t \to \infty$. See Section~\ref{app:prior_modelbased} in the supplementary materials for details on prior specification.

@@

\newpage
\textcolor{red}{Fix references!}
\bibliographystyle{plain}
\bibliography{covid-refs}

@@

For binary outcomes, we have
\begin{equation} \label{eq:weightedbinaryrho}
\rho_{\tilde I (X),Y} = \tilde \Delta \sqrt{\frac{\bar Y (1 - \bar Y)}{f (1-f) \text{E}(W_J \mid I_J = 1)^2 + f \text{Var}(W_J \mid I_J = 1)} }
\end{equation}

@@

\begin{figure}[!th]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig1.png}
 \caption{Symptom Likelihood Given Negative Test}
 \label{fig:symptomlik1_model2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
\includegraphics[width=.9\linewidth]{../figs/tvprop_symptom_alt_fig2.png}
 \caption{Symptom Likelihood Given Positive Test}
 \label{fig:symptomlik2_model2}
\end{subfigure}
\caption{Likelihood of COVID-19 contact}
\label{fig:symptomlik_model2}
\end{figure}

@@

\section{Model-based: Prior specification}
\label{app:prior_modelbased}
